Measuring Hash Trustworthiness via Collision Utility Metrics: Logical Cryptanalysis of MD4

### 2017-12-03 dsn-2018/bare_conf.tex
Extend Techniques section
Signed-off-by: Alexander Scheel <alexander.m.scheel@gmail.com>

We introduce a distance function, $\delta$, between collision classes by the
number of differences in intermediate rounds deltas. That is, given two
collision classes, $C_1, C_2 \in \mathcal{C}$:
\begin{align}
    \delta(C_1, C_2) = \big| \{ i : C_1[i] \neq C_2[i] \} \big|
\end{align}

@@

The second metric evaluates the density of the neighborhood of a collision
class. If a collision resides in a dense neighborhood, it provides more
possible collision classes to search for a second preimage, chosen prefix, or
other structure desired in a collision. If however, a collision class has no
neighbors, then it cannot be used to find other possible classes for other
input blocks.

@@

If a collision class satisfies many of these properties, then it is more
flexible and thus more likely to be used to target deployed systems. If,
however, a collision class does not satisfy these properties, it impact is
likely severely limited in scope, and may not provide useful information to
find other collision classes which have higher utility.

@@

We find justification for this metric in the existing literature on MD4:
H. Dobbertin's \cite{Dobbertin1998} and P. Kasselman's \cite{KasselmanMD4}
collision classes have distance $1$ under this metric. We extend this distance
function to include an \textit{absolute} distance, measuring the number of
intermediate rounds with non-zero differences. We denote this as $\delta(C_1)$.

@@

\subsection{Family Similarity}
We define a relation among families of collisions across rounds. Let $F_1$ and
$F_2$ be two different families of collisions. Then we say that $F_1$ and $F_2$
are \textit{similar}, and notate it $F_1 \lesssim F_2$, if $R(F_1) \leq R(F_2)$
and $F_1 \subseteq F_2$. Note that, when $R(F_1) = R(F_2)$, this is equivalent
to saying that $F_2$ is in some expansion neighborhood of $F_1$. Further, when
$R(F_1) < R(F_2)$ and $F_1 = F_2$, then we say that $F_2$ is the
\textit{trivial extension} of $F_1$.

We claim that the following statements are true:
\begin{enumerate}
    \item For every $F \in \mathcal{F}$, there exists $F'$ such that
        $F' \lesssim F$ and $R(F') + 4 = R(F)$.
    \item For every $F \in \mathcal{F}$, there exists $F'$ such that
        $F \lesssim F'$.
\end{enumerate}

\textbf{TODO} - Describe justification? Later sections?

\subsection{Class Similarity}

We claim that the aforementioned statements hold when considering individual
collision classes instead of families of collision classes, but with less
likelihood.

\textbf{TODO} - Describe technique applied neighborhood extensions

Give some reasoning for why it works, example graphs.

\subsection{Miscellaneous}

\textbf{TODO} - Describe chosen prefix, second preimage, ASCII, and other
tips and tricks for faster model runs.

\section{Empirical Results}
\subsection{Unique Differentials}

\textbf{TODO}
Refer to Table \ref{table:differentials} for the number of differential
paths.

>>>>

\subsection{Distance Metrics} \label{terminology:distance}

We introduce a distance function, $\delta$, between collision classes by the
number of differences in intermediate rounds deltas. That is, given two
collision classes, $C_1, C_2 \in \mathcal{C}$:
\begin{align}
    \delta(C_1, C_2) = \big| \{ i : C_1[i] \neq C_2[i] \} \big|
\end{align}
We extend this distance function to include an \textit{absolute} distance,
measuring the number of intermediate rounds with non-zero differences.
We notate this as $\delta(C_1)$.

@@

The second metric evaluates the density of the neighborhood of a collision
class. If a collision resides in a dense neighborhood, it provides more
possible collision classes to search for a second preimage, chosen prefix, or
other structure desired in a collision. If, however, a collision class has no
neighbors, then it cannot be used to find other possible classes for other
input blocks.

@@

If a collision class satisfies many of these properties, then it is more
flexible and thus more likely to be used to target deployed systems. If,
however, a collision class does not satisfy these properties, its impact is
likely severely limited in scope, and may not provide useful information to
find other collision classes which have higher utility.

@@

We find justification for the distance metric defined in Section
\ref{terminology:distance} in the existing literature on MD4:
H. Dobbertin's \cite{Dobbertin1998} and P. Kasselman's \cite{KasselmanMD4}
collision classes have distance $\delta(C_{Dobbertin}, C_{Kasselman}) = 1$
under this metric. The distance function allows us to find collision classes
based off of existing classes. By limiting ourselves to unit distances, we
maximally reuse an existing collision and limit the amount of work the SAT
solver needs to do. Further, rather than searching for any unit-distance
new collision class, we can easily split our search to look in each of the
48 rounds of MD4, allowing us to parallelize the search for neighbors.

@@

\subsection{Family Similarity}

We define a relation among families of collisions across rounds as follows.
Let $F_1$ and $F_2$ be two different families of collisions. Then we say that
$F_1$ and $F_2$ are \textit{similar}, and notate it $F_1 \lesssim F_2$, if
$R(F_1) \leq R(F_2)$ and $F_1 \subseteq F_2$. Note that, when
$R(F_1) = R(F_2)$, this is equivalent to saying that $F_2$ is in some
expansion neighborhood of $F_1$. Further, when $R(F_1) < R(F_2)$ and
$F_1 = F_2$, then we say that $F_2$ is the \textit{trivial extension} of
$F_1$.

We claim that the following statements are true for MD4:
\begin{enumerate}
    \item For every $F \in \mathcal{F}$, there exists $F'$ such that
        $F' \lesssim F$ and $R(F') + 4 = R(F)$.
    \item For every $F \in \mathcal{F}$, there exists $F'$ such that
        $F \lesssim F'$.
\end{enumerate}

From an intuitive standpoint, what this is says is that collisions have
witnesses for the existence in reduced round versions of the function, and
further, that it is sufficient to search reduced round versions to find
candidates for extension into additional rounds. This likely derives from
the iterative construction of MD4 and the particular choice of block schedule.

\textbf{TODO -- Empirical justification / graphical representation?}

\subsubsection{Collision Class Similarity}

We claim that the aforementioned statements hold when considering individual
collision classes instead of families of collision classes, but with less
likelihood. That is, we can transplant a $n$ round collision into a $n+4$
round and consider a neighborhood around it to find a new collision.

\subsubsection{Second Preimage Similarity}

If the models are constrained for a second preimage, we note that attack again
works, but requires searching wide and wider differential paths in small round
sizes. In particular, despite strong evidence for this pattern holding in
$n \leq 24$ rounds, outside of the collision family $\{0, 16\}$ in both 28 and
32 round MD4--which does not extend any further--we have been unable to find
any other differential path in those rounds by exhaustive brute force search
up to a differential path of length 8. Further searches have been done of
36, 40, 44, and 48 rounds up to a differential path of length 5 with a similar
lack of results. Thus we believe the general technique to hold, but the size
of the search space for a second preimage becomes prohibitively large, and
further, that differential paths become sufficiently long to prohibit easy
finding of second preimages.

\subsubsection{Chosen Prefix Similarity}

Unlike second preimage attacks, chosen prefix attacks have a higher likelihood
of working, and give better results. However, we have found that most chosen
prefixes work in the neighborhoods of existing collisions, and thus are not
convinced that inductively building a chosen prefix collision is a useful
technique by itself.

\subsection{Unjustified Performance Observations}

We share the general tips for working with models of MD4 from a SAT solver
performance perspective, without justification for results. Note that these
results may be limited in usefulness and may not be uniformly applicable.
\begin{itemize}
    \item Unless searching for a second preimage or chosen prefix with
        sufficient number of bits ($\geq 128$), we find models perform better
        leaving the initial state unconstrained.
    \item Adding ASCII block constraints significantly and negatively impact
        the performance of finding a collision.
    \item Keeping small search radius on neighborhoods involving families
        of collision classes is important for good performance. However,
        once sufficient witnesses have been found across rounds, missing
        families can be found by various filling-in techniques by applying
        the $\lesssim$ relation.
\end{itemize}

\section{Empirical Results:\\The Utility of Existing Collisions}

\textbf{TODO -- Introduction?}

\subsection{Unique Differentials}

Table \ref{table:differentials} states the
